Using SMT engine to generate Symbolic Automata\thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2018-01-13 2017-Xudong-SMT/FM-2018-extended.tex

@@

@@

\section{Implementation}
\label{section:implementation}
The VerCors platform uses pNets as the intermediate language
for some high-level language or graphical formalism to be translated
into both input for a model-checker and for generating executable code
automatically \cite{HKM-FASE16}.
We have extended the pNet API in VerCors to deal with open pNets, and
also to specify the structure of action algebras.

@@

%\begin{algorithm}
%\caption{Open Automaton Generation}
%\begin{algorithmic}[1]
%
%\State Initialize sets $U$, $E$ for unexplored/explored global states
%$U$, $L$ for result OTs;
%\While{!isEmpty(U)}
%	\State Chose a global state $S$ in $U$; recursively applying Tr1 or Tr2 on the root node;
%	\If{Subnet is a pLTS}
%                \State Get the local state of the pLTS $s$ from $S$;
%		\For{each transition $tr$ starting from $s$} Apply Tr1
%			\State Build the corresponding
%                        $OT$; store $OT$ in $L$; 
%                 \EndFor
%	\EndIf
%	\If{Subnet is a pNet}
%		\For{each $sv \in SV$} Apply Tr2
%			\State Invoke $Combining()$ to generate the
%                        combination of sub-transitions, $Matching()$
%                        to match the combination with the $sv$; Store the result OTs in $L$; 
%		\EndFor
%	\EndIf
%	\For{each target global state $T$ from $OT \in L$}
%		\IfThen{$(!U.contains(T))\ \&\&\ (!E.contains(T))$}{Add $T$ into the $U$;}
%	\EndFor
%\EndWhile
%\State Prune the open transitions in $L$ using the SMT solver;
%\State \Return $L$;
%
%\end{algorithmic}  
%\end{algorithm}

@@

\subsection{The Generation of Open Automata}

\QIN{
We have already have the sketch of computing the open automata \cite{???}.
While in our implementation we does not build explicitly a proof tree for every open transition.
Instead, we repeat applying the Tr1 or Tr2 to generate the sub-transitions.
Tr1 is applied on the leaf (pLTS) simply take the pLTS transitions through the given start state and add the predicate in pNet to generate the open transition.
When applying Tr2 we use two methods, combining and matching, to generate the new predicate of the open transitions in a hierarchical manner
as in this case the composition the subnets brings more constraints for synchronizing behaviors. 
Beside the new generated predicate, the predicates from the subnets are also added as a conjunction with those new predicates.
}

@@

%From these, we compute
%the open automaton of an open pNet. We explain here our main
%implementation choices.
%:
%\begin{itemize}
%  \item Steps 2a-2b are merged, applying Tr2 from the
%premisses OTs from subnets to generate OTs without
%explicitely constructing the deduction tree.
%  \item We define a general naming schema for implementing the fresh 
%and clone functions with more structure than a simple global name
%generator. It ensures unicity, but builds names
%readable enough for debugging purpose.
%  \item Besides that, we also have designed a management of state variables and
%    assignments throughout the whole computation.
%%%   \item Satisfiability check is done only at
%%% top level of the open-transition construction. An alternative would be
%%% to submit the satisfiability check to the SMT solver at each level of
%%% the hierarchical OT construction, potentially reducing the overall number of
%%% combinations. But the submission to the SMT engine is costly, and more
%%% complexity analysis is required before deciding if this would be
%%% worthwhile. 
%\end{itemize}
%
%In the following sections, we detail:
%\begin{itemize}
%\item The structure of fresh variable names.
%\item The details of the implementation for step 2a-2b, refined into:
%  \subitem combining all possibilities of subnets activity/inactivity,
%  \subitem matching each combination with synchronization vectors.
%\item Management of local variables of pLTSs.
%\item Pruning unsatisfiable OTs using the Z3 SMT solver.
%\end{itemize}

%\TODO{I have not yet been in details through the next section, but we
%  need to shorten them...}

% \TODO{Xudong: move here a (new version) of the alogorithm loplevel sketch}
% \QIN{[Move to the head of Section 5.]}

%% \QIN{
%% The detail about the application of the Tr1 and Tr2 have not considered in the sketch, but it is an  
%% important part in the implementation of open transition generation. 
%% Each time we apply Tr1, there is not too much predicate terms, so do the other parameters, needed 
%% to be generated. 
%% When it comes to applying Tr2, the computation becomes complicated because it doing some composition of the subnets.
%% }

%\paragraph{Combining and Matching using TR2}
%Our implementation does not build explicitly the proof trees for each
%open transition, but instead : [Combining] enumerate all possible combinations of
%pLTS transition (using TR1) and each synchronization vector choice
%(TR2); [Matching] build the corresponding transition, and in particular
%its predicate, in a hierarchical manner.
%Each time we apply Tr2, we build a predicate as a conjunction of two
%parts. The first part, which is the simpler one, composes the
%predicate from the 
%subnets and the guard from transitions in sub-pLTS or synchronization
%vectors. The second part is the conjunction of equations generated
%during matching the subnets' behaviors with the synchronization
%vectors.

%Note that it is simple to collect required terms in the first
%part while in the second part we need to find out all the possible
%conjunction forms. We present here a algorithm divided in two
%step. Combining enumerates all the possible combinations of the
%working status. Matching synchronizes the subnets according to
%synchronization vectors to generate possible predicates.

%Combining is to generate all the possible matchings of the subnets' OTs (open transitions). The result list of the matched tuples called *combination*.
% 1. Initialize the result list of OT tuples.
% 2. Get one list of OTs. It's same as get OTs of one subnet.
% 3. Add *null* into the list, it means this subnet does not work this time.
% 4. Combine the rest subnets' OTs. Get the combination of the rest OTs.
% 5. If there are OTs of the chosen list not be chosen, choose one OT.
%      1. If there are tuples of the combination not be chosen, choose one tuple.
%          1. Add the chosen OT into the tuple.
%          2. Add the tuple into the result list, then do the recursion.
% 6. Return the result list.

@@

% Mixing is to match the combination of the subnets' open transitions, hole behaviors and result action with the SV (synchronization vector) to generate the possible predicates of the OT (open transition). At the same time, other parameters of the OT (global state, subnets' transitions, result action...) are also generated. Then the possible OTs of the pNet can be achieved. 
% 1. Initialize the combination through the method *combineOpenTransitions( )*.
% 2. Initialize the result list.
% 3. Get the number of subnets and holes. They are used to decide how many element of the SV should be matched with them.
% 4. For all tuples of the combination.
%      1. Invoke the *clone( )* method to generate the fresh SV.
%      2. Match the elements of the tuple with the elements of the SV. If there is a *null* matched with a *not null* action, filter this matching. Add the result expression and the predicate of the subnets into the predicate. Generate the target global state correspondingly.
%      3. Match the hole behaviors with the elements of the SV. The hole behavior could be *null*. Add the result expression into the predicate.
%      4. Match the result action with the result action of SV. Add the result expression into the predicate.
%      5. Add the predicate into the OT. Store the OT in the result list.  
% 5. Return the result list of all the possible OTs.

%We now come to the matching using combinations from the previous step, together with hole behaviors and the result action, to match with the synchronization vectors. 
%
%Algorithm 2 shows the algorithm matches combination $L_C$, hole
%behaviors $B$ with synchronization vectors $SV$. Remember variables
%inside a synchronization vector are consider local to the vector; so every time we choose a synchronization vector $sv$ from $SV$ for matching, we clone it and use the cloned one $sv'$. We use the definition declared before to construct all the possible predicates but the algorithm has not checked the correctness of these predicates. 
%
%There is one mismatch case we can easily detect here, that is its
%working status:
%At any time an action $e_1$ matching with an element $e_2$ from
%synchronization vector, the pair $(e_1, e_2)$ must be checked: if both
%$e_1$ and $e_2$ are $null$ or $not\ null$, then it continues to
%generate a new term of predicate. Otherwise, it is intuitive that the
%working status coming from the subnet $e_1$ is different from what
%asked by the synchronization vector $e_2$, then this case is filtered
%out.

%\begin{algorithm}
%\caption{Matching}
%\begin{algorithmic}[1]
%%\Require 
%%The combination of subnets' open transitions $L_C$;
%%The behaviors of the holes $B$;
%%The set of synchronization vectors $SV$.
%%\Ensure 
%%The list of all the possible open transitions generated $L_{ot}$.
%\State Initial an empty result list $L_{ot}$;
%\For{each $C \in L_C$}
%	\For{each $sv \in SV$}
%		\State Clone the $sv$, the result is $sv'$;
%		\State Generate the fresh result action $ra$;
%		\State Combine $C, B, ra$ as a new vector $v$;
%		\For{each $(e_1 \in v )\&\& (e_2 \in sv')$}
%			\If{$(e_1\ is\ null\ \&\&\ e_2\ is\ not\ null)\ ||\ (e_1\ is\ not\ null\ \&\&\ e_2\ is\ null)$} 
%			\State Skip;
%			\Else ~ Add $(e_1 = e_2)$ into the $Pred$;
%			\EndIf
%		\EndFor
%		\State Generate the result open transition $ot$ with $Pred$;
%		\State Add the $ot$ into $L_{ot}$;
%	\EndFor
%\EndFor 
%\State \Return $L_{ot}$;
%
%\end{algorithmic} 
%\end{algorithm}

@@

\QIN{
$$Matching(SV_k, C_n, Hole, v) = 
\begin{array}{l}
\forall {(a'_i)}^{i\in I},
{(b'_j)}^{j\in J},v'.
\\
v_i=a'_i\land \, b_j=b'_j \land v=v' \land G_k
\end{array}$$
which is exactly the new part of predicate for one possible result.
}
%Algorithm 3 shows the algorithm matches
%combination $L_C$, hole behaviors $B$ with synchronization vectors
%$SV$ according to the definition of the predicate declared before. We
%clone the $sv$ using the fresh function to generate a new $sv'$ with
%fresh variables.The mismatch on working status is easy to detect so we
%filter them in the algorithm through checking the matching elements
%$e_1$ and $e_2$. 
%Then we build the resulting open transition. Satisfiability of the
%predicate is not checked at this level, but only later, at the
%toplevel of the pNet.

\paragraph{Filtering:} 
\QIN{There is already able to filter some wrong open transitions during the matching.
We can ensure that the synchronization vector is not matched with the subnets if there is only one side not involved.
Then we can remove the transitions from the result of matching according to the condition $\gamma$ where
$\gamma = \bigcup_{i\in I}(v_i\in\eta\land a'_i\in\eta)\lor(v_i\in\eta\land a'_i\in\eta) .$
}

@@

\subsection{Transition to SMTlib}
In order to submit satisfiability problems to Z3 (for the predicates
in open transitions), we need to generate SMTlib programs, from the
pNet Algebra presentation and predicates.
More precisely, we need to translate to SMTlib:
\begin{itemize}
  \item the presentation of the action algebra
    (sorts and operators) that is defined for a given language (process calculus, or high level programming language),
    \item for a given pNet, the set of local constants (actions or auxiliary data) that are used in the pLTSs,
    \item for each open transition: the declaration of variables, and
      the predicate (including action expressions).
\end{itemize}

@@

\begin{example}
  \begin{itemize}
  \item $Sorts_{CCS} = \{Action, Channel, Data, Int, Bool\}$
  \item $Constrs_{CCS} =\\
    \texttt{Emit}: 2, \{(Chan\_E:Channel),(Value\_E:Data)\}->Action\\
    \texttt{Receive}: 2, \{(Chan\_R:Channel),(Value\_R:Data)\}->Action\\
    \texttt{Tau}: 0, \{\}-> Action\\
    \texttt{... and all predefined operators}$
  \end{itemize}
\end{example}

\begin{lstlisting}
AlgebraSort Action = new AlgebraSortImpl("Action");
AlgebraSort Channel = new AlgebraSortImpl("Channel");
AlgebraSort Data = new AlgebraSortImpl("Data");
Action.addConstructor("Emit", 
	{"Chan_E", "Value_E"}, {Channel, Data});
Action.addConstructor("Receive", 
	{"Chan_E", "Value_E"}, {Channel, Data});
Action.addConstructor("Tau");
\end{lstlisting}

\TODO{[DONE][Xudong]: Give an example of defining $Tau$ and $Emit$ using the (new
  version of) the API}

\subsubsection{Expressions}
Once an algebra presentation is defined, we can construct expressions
of the term algebra, using variables and operators:

@@

The purpose of this section is to define static semantic notions that
will guarantee that the translation to the SMT language will be
correct, i.e. will not yield errors at runtime. This includes
well-formness (all sorts, operators, variables are defined, and
expressions respect the arity of operators), and typing rules.

@@

\smallskip
\ERIC{My plan here is to provide the translation principles as some
  high-level pseudo code, together with precise definition of the
  translation functions of a presentation+environment, and of a
  predicate (including expressions)}

@@

\subsubsection{Presentation Translation}
We define here the translation of the algebra presentation, extended
with the constant operators collected from the pNet. It produces the
declaration of sorts (excepted Bool and Int) with their constructors
and selectors as \texttt{declare-datatypes}, and the declaration of
other operators as \texttt{declare-function} constructs.
For sorts, we must distinguish the case of mutually defined sorts
(e.g. edges and vertices in a graph), that must be declared within a
single \texttt{declare-datatypes} construct. For this we define:

@@

\TODO{[Xudong: this is certainly obsolete?] At the same time, we have
  several internal lists (actually hash 
  maps) \texttt{exprs}, \texttt{funcDecls}, \texttt{sortDatatypes} to
  store the generated Z3 objects with their names for later proofs.}

@@

\noindent
\begin{tabular}{p{5cm}p{5cm}p{2.5cm}}
		\hline\specialrule{0em}{3pt}{3pt}
		\multicolumn{2}{l}{Variables}

@@

\subsection{Simplification of Predicates}
\TODO{[Eric]: Motivate, and refs  -- reduce the section here to approx
  half a page} 
\QIN{
We apply several structural rules to generate the predicates restricting the composition of the subnets by synchronization vectors.
The predicates may contain some redundancy if there are variables in middle level are matching other variables both in lower and higher level at the same time. 
}

@@

\TODO{Update: we have 11 satisfiable OTs here}
Our tool builds 30 open
transitions, out of which 27 are detected unsatisfiable by Z3, we
get only 3 results left; the resulting open automaton is shown in
Fig. ~\ref{schema:lotos-result}, together with its open transitions.

>>>>

\TODO{Eric: merge term algebra, algebra pres, and add a short example (BIP)}

@@

\TODO{Simon: or may be better, present BIP architectures and the use-case \_BEFORE\_ The pNets definition, and define the encoding here}

@@

\section{Generation of Open Automata}
\label{section:implementation}

\TODO{Eric: add a header, reorganise the parts of the algorithm}

@@

\TODO{Eric: rewrite}
We have already have the sketch of computing the open automata \cite{???}.
While in our implementation we does not build explicitly a proof tree for every open transition.
Instead, we repeat applying the Tr1 or Tr2 to generate the sub-transitions.
Tr1 is applied on the leaf (pLTS) simply take the pLTS transitions through the given start state and add the predicate in pNet to generate the open transition.
When applying Tr2 we use two methods, combining and matching, to generate the new predicate of the open transitions in a hierarchical manner
as in this case the composition the subnets brings more constraints for synchronizing behaviors. 
Beside the new generated predicate, the predicates from the subnets are also added as a conjunction with those new predicates.

@@

@@

@@

@@

\QIN{
$$Matching(SV_k, C_n, Hole, v) = 
\begin{array}{l}
\forall {(a'_i)}^{i\in I},
{(b'_j)}^{j\in J},v'.
\\
v_i=a'_i\land \, b_j=b'_j \land v=v' \land G_k
\end{array}$$
which is exactly the new part of predicate for one possible result.
}

\paragraph{Filtering:}
\TODO{Eric: 3 levels of filtering, explain our choice}

@@

\subsection{Translation to SMTlib}

\TODO{Eric: Replace by (excerpts of) the formal version, from appendix, and reduce to less than one page}

As seen in section
\ref{section:pnets}, the \emph{term algebra} used to express the
action expressions and data parameters is left open in
the pNet definition.

\subsection{Simplification of Predicates}
\ERIC{: motivated, add refs, and an abstract presentation of stuff from pages 19-20}

\TODO{Xudong: Update: we have 11 satisfiable OTs here}
Our tool builds 30 open
transitions, out of which 27 are detected unsatisfiable by Z3, we
get only 3 results left; the resulting open automaton is shown in
Fig. ~\ref{schema:lotos-result}, together with its open transitions.

\section{More details on the generation algorithm}
\subsection{Algebra specification}
\subsection{Term algebra: type system and static semantics}
Our models rely on a notion of parameterized actions, that are
symbolic expressions using data types and variables. As our model aims
at encoding the low-level behavior of possibly very different
programming languages, we do not want to impose one specific algebra
for denoting actions, nor any specific communication mechanism. So we
leave unspecified the constructors of the algebra that will allow building
expressions and actions. Moreover, we use a generic {\em action interaction}
mechanism, based on unification between two or more action
expressions. This will be used in the semantics of synchronization
vectors to express various kinds of communication or synchronization mechanisms.

\subsection{Fresh variables}

For example, in the running example, the first possible behavior of
the hole, it doesn't have a name, only have the prefix ``hb''. The
hole is the second subnets of the root node so its ``tree index'' is
``12''. The fresh variable name is ``$\OTvar{:hb:12:1}$''. 
\subsection{Variable managment}
The variables in each synchronisation vector are considered local: for a given pNet expression, we must have fresh local variables for each occurrence of a vector (= each time we instantiate rule Tr2). Similarly the state variables of each copy of a given pLTS in the system, must be distinct, and those created for each application of Tr2 have to be fresh and all distinct. This will be implemented within the open-automaton generation algorithm, e.g. using name generation using a global counter as a suffix.
\subsection{Filtering the behaviour combinations}
\QIN{There is already able to filter some wrong open transitions during the matching.
We can ensure that the synchronization vector is not matched with the subnets if there is only one side not involved.
Then we can remove the transitions from the result of matching according to the condition $\gamma$ where
$\gamma = \bigcup_{i\in I}(v_i\in\eta\land a'_i\in\eta)\lor(v_i\in\eta\land a'_i\in\eta) .$
}
\subsection{Translation to SMTlib input language}
In order to submit satisfiability problems to Z3 (for the predicates
in open transitions), we need to generate SMTlib programs, from the
pNet Algebra presentation and predicates.
More precisely, we need to translate to SMTlib:
\begin{itemize}
  \item the presentation of the action algebra
    (sorts and operators) that is defined for a given language (process calculus, or high level programming language),
    \item for a given pNet, the set of local constants (actions or auxiliary data) that are used in the pLTSs,
    \item for each open transition: the declaration of variables, and
      the predicate (including action expressions).
\end{itemize}

@@

\begin{example}
  \begin{itemize}
    \item[]
    \item $Sorts_{CCS} = \{Action, Channel, Data, Int, Bool\}$
    \item $Constrs_{CCS} =\\
    \texttt{Emit}: 2, \{(Chan\_E:Channel),(Value\_E:Data)\}->Action\\
    \texttt{Receive}: 2, \{(Chan\_R:Channel),(Value\_R:Data)\}->Action\\
    \texttt{Tau}: 0, \{\}-> Action\\
    \texttt{... and all predefined operators}$
  \end{itemize}
\end{example}

Then in pNet API these operators or constants (seemed as non-parameter operators) can be declared as:
\begin{lstlisting}[basicstyle=\footnotesize\ttfamily, language=java, frame=single]
AlgebraSort Action = new AlgebraSortImpl("Action");
AlgebraSort Channel = new AlgebraSortImpl("Channel");
AlgebraSort Data = new AlgebraSortImpl("Data");
Action.addConstructor("Emit", 
	{"Chan_E", "Value_E"}, {Channel, Data});
Action.addConstructor("Receive", 
	{"Chan_E", "Value_E"}, {Channel, Data});
Action.addConstructor("Tau");
\end{lstlisting}

\paragraph{Expressions}
Once an algebra presentation is defined, we can construct expressions
of the term algebra, using variables and operators:

@@

The purpose of this section is to define static semantic notions that
will guarantee that the translation to the SMT language will be
correct, i.e. will not yield errors at runtime. This includes
well-formness (all sorts, operators, variables are defined, and
expressions respect the arity of operators), and typing rules.
\TODO{Eric: refer in the appendix header}

@@

\smallskip
%% \ERIC{My plan here is to provide the translation principles as some
%%   high-level pseudo code, together with precise definition of the
%%   translation functions of a presentation+environment, and of a
%%   predicate (including expressions)}

@@

\subsubsection{Presentation Translation} \TODO{Duplicate ?}
We define here the translation of the algebra presentation, extended
with the constant operators collected from the pNet. It produces the
declaration of sorts (excepted Bool and Int) with their constructors
and selectors as \texttt{declare-datatypes}, and the declaration of
other operators as \texttt{declare-function} constructs.
For sorts, we must distinguish the case of mutually defined sorts
(e.g. edges and vertices in a graph), that must be declared within a
single \texttt{declare-datatypes} construct. For this we define:

@@

@@

\noindent
\TODO{This is incomplete... Xudong: Conjunctive normal form, translation function for boolean expressions, + assignments}

\begin{tabular}{p{5cm}p{5cm}p{2.5cm}}
		\hline\specialrule{0em}{3pt}{3pt}
		\multicolumn{2}{l}{Variables}

@@

\subsection{Simplification of Predicates}
\TODO{[Eric]: Motivate, and refs  -- reduce the section here to approx
  half a page (from the next two pages...)}

\subsection{Elimination of intermediate variables}

\QIN{
We apply several structural rules to generate the predicates restricting the composition of the subnets by synchronization vectors.
The predicates may contain some redundancy if there are variables in middle level are matching other variables both in lower and higher level at the same time. 
}

@@
